jstor: Parsing XML - Lessons From Writing a Package

### 2018-09-30 draft.Rmd
expand draft

---
title: 'jstor: Parsing XML - Lessons From Writing a Package'
author: "Thomas Klebel"
date: "25 9 2018"
output: html_document
---

@@

# Step 1: Hacking Away
Back in March 2017, I was starting out as a student in a research project
concerned with the scientific elite within sociology and economics. The project
had many goals, but writing an R package was not one of them. At the beginning,
I was presented with a dataset which was huge, at least for my terms: 
around 30GB of data, half of which was text, the other half 500,000 .xml-files.
The dataset was incredible in its depth: we basically sat on all articles
from JSTOR which belonged to the topics "sociology" and "economics". To repeat:
all articles that JSTOR has on those topics for all years.

@@

```r
file_path %>% 
  read_xml() %>% 
  as_list() %>% 
  find_meta()
```

As can be expected when parsing deeply nested lists which do not always have
the same structure, this quickly escalated into
more and more complex or sometimes quite ridicoulus functions like:

@@

At some point though, I started benchmarking my solutions. I tried out 
`mclapply` and it doubled the speed of execution on my machine, but it was still
very slow: Parsing the content *after* importing and transforming it into a list
would take 7.2 seconds for 3000 files. For 200,000 files of sociological 
articles this would amount to `r 7.2*200000/3000/60` hours. And importing and
converting to a list took a lot of time too. Although I didn't thought, I would
need to import the files many times (which I had to do later to update results
after we got new versions with better data), I decided that this was too slow,
and I looked for options.

The first idea which I had was simply to scale computing power: I had a faster
machine at home, and planned to read the data into R at work, save it as .rds,
and then process it at home on the faster machine (for some reason I didn't want
to carry the original data home). Apperently, this is not as easy as it sounds.
If you parse a file with `xml2::read_xml()` you don't get the content of the
file, but only an object that points to the file, and you can then extract 
content with `xml_child()` and other functions.
But you cannot use `readr::write_rds` to save the file, instead you need
`xml2::xml_serialize()` and `xml2::xml_unserialize()` and they require an open
connection, and so on. I can tell you that, since Jim Hester kindly responded to [my
question on StackOverflow](https://stackoverflow.com/questions/44070577/write-xml-object-to-disk). 
He had a further hint:

@@

And this is what I did. For the fun of it, I benchmarked two versions: my 
function before re-writing to using XPATH expressions, and the function today.
The current version is a lot safer and handles many edge-cases, but it is still
around 5 times faster. Furthermore, it depends less on how complex the original
file is, and more on how much information is being read. With the new version
of the function, around 30% of the runtime is spent on reading the file, while 
with the old version it was around 85% of the time.

@@

The work on re-writing progressed quickly. The first step was to simply extract the
identifier for the journal within the article it was published:

@@

Another thing I took up
during this re-write was regularly benchmarking my functions. Since I lacked 
(and still lack) the technical expertise to know which approach would be faster,
I had to benchmark possibilities to figure it out:

@@

microbenchmark::microbenchmark(fun1(), fun2(), fun3(), times = 1000)
# Unit: microseconds
# expr     min       lq     mean   median       uq      max neval
# fun1() 267.684 290.5885 322.8992 301.4325 318.9060 2705.459  1000
# fun2() 233.723 250.3840 287.1126 263.1725 273.2180 2716.416  1000
# fun3() 686.542 725.0925 822.2103 763.4240 798.1155 3534.849  1000
```
Although the unit is microseconds here, the difference between `fun2` and 
`fun3` is quite substantial, if you want to do it 500,000 times: 4.5 minutes. 
I'd guess that `fun2` is faster, since it only calls two functions, and not 
four, and because `xml_text()` probably converts the content directly to 
character in C++, which is faster than calling `as.character` afterwards. But
I am still a *social* scientist, so I wouldn't bet on that explanation.

After rewriting and expanding the functionality, I was still not happy, however,
since the code was slower than I thought it should be. At this point,
I dug deeper again, using the package `profvis` to find bottlenecks. Oddly
enough, I had introduced a bottleneck right at the beginning: I was using
`data_frame` (which is equivalent to `tibble()`) to create the object my 
function would return. Unfortunately, `tibble` (and `data.frame` as well) are
quite complex functions. They do type
checking and other things, and if you do this repeatedly, it quickly adds up.

@@

But how did I find the functions responsible for the delay? I used the package
[profvis](https://rstudio.github.io/profvis/index.html)
a lot during this progress. The package measures time spent line by
line, which let's you identify bottlenecks in your code.

@@

A simple solution is to rerun the function many times. Note, however, that this
may lead to varying results. The function is not measured 500 times, the result
being the mean or median (like in `microbenchmark`), but it is simply the 
aggregate of running the function 500 times. This can vary quite a bit. All in
all, tough, I found it still useful to judge, if any part of the code is orders 
of magnitude slower than the rest.

@@

```{r}
profvis::profvis({
  old_approach <- function() {
    authors <- get_authors()

    res <- tibble::data_frame(id = "1",
                              authors = list(authors))

@@

  purrr::rerun(500, old_approach())
})
```

We can see that the call to `tidyr::unnest` takes up a lot of time, thus
it would be good to get rid of it. The new approach does not need unnesting, and
is around three times faster:

@@

  new_approach <- function() {
    authors <- get_authors()

    data.frame(id = "1",
               authors = authors)
  }

  purrr::rerun(500, new_approach())
})
```

By assessing the efficiency of functions repeatedly and optimising several parts,
I was able to trim down execution time considerably overall. For 25,000 files,
which is the maximum amount of files you can receive at one time through the
interface of JSTOR/DfR, execution time is slightly under 4 minutes, or 2 minutes
if executed in parallel, at least on my moderately fast MacBook Pro.

# Lesson learned:

- community is important

# Lesson four: what could I have done differently?
Decision for API: each function is self-contained. If I had extracted the parts
for import somehow, it would be faster to apply different functions (like for
metada and references and other stuff).

>>>>

---
title: 'jstor: Parsing XML - Lessons From Writing a Package'
author: "Thomas Klebel"
date: "30 9 2018"
output: html_document
---

@@

Before taking on the look back into the past, I want to give you some basic
information. JSTOR is a large archive for scientific texts, mainly known for 
their coverage of journal articles, although they recently added book chapters
and other sources as well. They make all of their content available for 
researchers to do text mining, citation analysis, and everything else you could
come up using metadata and data about the content of the articles. The package
I wrote, `jstor`, provides functions to import the metadata, and it has a few
helper functions for common cleaning tasks. Everything you need to know about 
how to use it is on the [package website](https://ropensci.github.io/jstor/).
You will find three vignettes, with a general introduction, examples on how to
import many files at once, and a few examples of known quirks of JSTORs data.
There is also a lengthy case study that also shows how to combine metadata and
data about the content of the articles. But for now let us turn back the clock
to follow along my journey of developing the package.

# Step 1: Hacking Away
Back in March 2017, I was starting out as a MA-student of
sociology in a research project
concerned with the scientific elite within sociology and economics. The project
had many goals, but writing an R package was not one of them. At the beginning
of my engagement,
I was presented with a dataset which was huge, at least for my terms: 
around 30GB of data, half of which was text, the other half 500,000 `.xml`-files.
The dataset was incredible in its depth: we basically sat on all articles
from JSTOR which belonged to the topics "sociology" and "economics". To repeat:
all articles that JSTOR has on those topics for all years.

@@

```r
# one file
file_path %>% 
  read_xml() %>% 
  as_list() %>% 
  find_meta()

# many files
file_paths %>% 
  map(read_xml) %>% 
  map(as_list) %>% 
  map(find_meta)
```

As can be expected when parsing deeply nested lists which do not always have
the same structure, this quickly escalated into
more and more complex and sometimes quite ridicoulus functions:

@@

At some point though, I started benchmarking my solutions. I tried out 
`mclapply` and it doubled the speed of execution on my machine, but it was still
very slow: Parsing the content *after* importing and transforming it into a list
took 7.2 seconds for 3000 files. For 200,000 files of sociological 
articles this would amount to `r 7.2*200000/3000/60` hours. And importing and
converting to a list took a lot of time too. Although I didn't think at that 
time, that I would
need to import the files more than once or at most twice
(which I had to do later to update results
after we got new versions with better data), I decided that this was too slow,
and I looked for options.

The first idea which I had was simply to scale computing power: I had a faster
machine at home, and planned to read the data into R at work, save it as `.rds`,
and then process it at home on the faster machine (for some reason I didn't want
to carry the original data home). Apparently, this is not as easy as it sounds.
If you parse a file with `xml2::read_xml()` you don't get the content of the
file, but only an object that points to the file, and you can then extract 
content with `xml_child()` and other functions.
But you cannot use `readr::write_rds` to save the file, instead you need
`xml2::xml_serialize()` and `xml2::xml_unserialize()` and they require an open
connection, and so on. I can tell you that, since Jim Hester kindly responded to [my
question on StackOverflow](https://stackoverflow.com/questions/44070577/write-xml-object-to-disk). 
He had a further hint:

@@

And this is what I did. For the fun of it, I benchmarked two versions: my 
function before re-writing to using XPATH expressions, and the function today.
The current version is a lot safer and handles many edge-cases, but it is still
around 5 times faster. Furthermore, it depends less on how complex the original
file is, and more on how much information is being read. With the new version
of the function, around 30% of the time is spent on reading the file, while 
with the old version it was around 85% of the time.

@@

The work on re-writing progressed quickly. The first step was to simply extract
the identifier for the journal within which the article was published:

@@

Another thing I took up
during this re-write was regularly benchmarking my functions. Since I lacked 
(and still lack) the technical expertise to know which approach would be faster,
I had to benchmark possibilities to figure it out. The following benchmark 
compares three versions of extracting the text from the node named `volume`.
Especially the first two seemed very similar to me and it was not at all 
obvious, which one would be faster.

@@

microbenchmark::microbenchmark(fun1(), fun2(), fun3(), times = 1000)
# Unit: microseconds
# expr     min       lq     mean   median       uq      max neval
# fun1() 267.684 290.5885 322.8992 301.4325 318.9060 2705.459  1000
# fun2() 233.723 250.3840 287.1126 263.1725 273.2180 2716.416  1000
# fun3() 686.542 725.0925 822.2103 763.4240 798.1155 3534.849  1000
```
Although the unit is microseconds here, the difference between `fun2` and 
`fun3` is quite substantial, if you want to do it 500,000 times: 4.5 minutes. If
you have to extract elements often (which I was doing), than this quickly adds 
up. What is the reason for this big difference?
I'd guess that `fun2` is fastest since it only calls two functions, and not 
four, and because `xml_text()` probably converts the content directly to 
character in C++, which is faster than calling `as.character` afterwards. But
I am still a *social* scientist, so I wouldn't bet on that explanation.

After rewriting and expanding the functionality, I was still not happy, however,
since the code was slower than I thought it should be. At this point,
I dug deeper again, using the package
[profvis](https://rstudio.github.io/profvis/) to find bottlenecks. Oddly
enough, I had introduced a bottleneck right at the beginning: I was using
`data_frame` (which is equivalent to `tibble()`) to create the object my 
function would return. Unfortunately, `tibble` (and `data.frame` as well) are
quite complex functions. They do type
checking and other things, and if you do this repeatedly, it quickly adds up and
is not very smart in general, since I know exactly what kind of data to expect
(if I wrote the rest of my functions appropriately).

@@

@@

A simple solution is to rerun the function many times. Note, however, that this
does not increase realiabity and reproducibility of the results. The function is
not measured 500 times, the result
being the mean or median (like in `microbenchmark`), but it is simply the 
aggregate of running the function 500 times. This can vary quite a bit. All in
all, tough, I found it still useful to judge if any part of the code is orders 
of magnitude slower than the rest.

@@

  old_approach <- function(authors) {
    res <- tibble::data_frame(id = "1",
                              authors = list(authors))

@@

  authors <- get_authors()
  purrr::rerun(500, old_approach(authors))
})
```

We can see that the call to `tidyr::unnest` takes up a lot of time, thus
it would be good to get rid of it. The new approach does not need unnesting, and
is roughly 17 times faster:

@@

  new_approach <- function(authors) {
    data.frame(id = "1",
               authors = authors)
  }

  authors <- get_authors()
  purrr::rerun(500, new_approach(authors))})
```

By assessing the efficiency of functions repeatedly and optimising several
parts,
I was able to trim down execution time considerably overall. For 25,000 files,
which is the maximum amount of files you can receive at one time through the
standard interface of JSTOR/DfR, execution time is slightly under 4 minutes, or
2 minutes if executed in parallel, at least on my moderately fast MacBook Pro.

# Lessons learned:

- community is important
- what could I have done differently?
Decision for API: each function is self-contained. If I had extracted the parts
for import somehow, it would be faster to apply different functions (like for
metada and references and other stuff).
